---
title: "Comparing networks"
author: "Nikolas Gestrich, Juan-Marco Puerta-Schardt"
date: "6/8/2021"
output: html_document
---
```{r libraries}
# Here are the libraries used in this example
library(tidyverse)
library(tidygraph)
library(igraph)
library(ggraph)
```

This document will show some basic ways in which networks can be compared. For archaeologists, this might be used to compare networks made on the same set of sites but whose edges are made on different materials, or for comparing different regions, or to assess the change of a network through time. The latter is what we will demonstrate here, using the Michelsberg data from the preceding instructions.

# Making the networks
The Michelsberg dataset contains a number of different periods. 
```{r check Michelsberg phases}
#this command will give us a list of periods and how many entries there are of each
Michelsberg %>% count(mbk_phase)
```
We can see that there are a lot of in between phases or uncertain attributions, and that there is quite a difference in the number of entries. In this case, we will artificially collapse these groups to make the numbers even, though we would need to justify this based on external arguments if this was real research work. As it is, we will place I together, with I/II, remove II/III, group III/IV with IV, and remove IV/V.
```{r adjust periods}
mbkphases <- Michelsberg %>%        #make a new dataset with modifications
                                    #adding the old phases to a new, simplified, phase column
  mutate(phase = case_when(mbk_phase == "I/II"| mbk_phase == "I" ~ 1,
                           mbk_phase == "II" ~ 2,
                           mbk_phase == "III" ~ 3,
                           mbk_phase == "III/IV"| mbk_phase == "IV" ~ 4,
                           TRUE ~ NA_real_))

# Look at the new phases
mbkphases %>% count(phase)
```
We now have four phases which have a tolerable number of observations in each. We will now make a network for each of them using the commands we learnt in the "Preparing data" file.
```{r make phase networks}
# to save us from typing too much, we will make a little function
select.phase <- function(p){
  mbkphases %>% 
    filter(phase == p) %>% # select only relevant phase
    select(5:39)  # select all columns that contain information on the amount of the different pottery types
}

# now apply this to the phases
mbk1 <- select.phase(1)
mbk2 <- select.phase(2)
mbk3 <- select.phase(3)
mbk4 <- select.phase(4)

#From these four datasets, we will now make edgelists
mbk1edges <- make.edgelist(mbk1)
mbk2edges <- make.edgelist(mbk2)
mbk3edges <- make.edgelist(mbk3)
mbk4edges <- make.edgelist(mbk4)

#The nodelist can be the same for all phases, we will trim the extraneous nodes later
mbknodes <- Michelsberg %>%
  rownames_to_column(var = "site_feature") %>%  # make the rownames into a separate column
  select(site_feature, site_name, feature_nr, x_utm32n, y_utm32n) %>%   # select the columns we want
  rename(lon = x_utm32n, lat = y_utm32n, id = site_feature) #change some of the more unwieldy names

#Now, we will create four graphs
mbk1net <- tbl_graph(nodes = mbknodes, edges = mbk1edges, directed = FALSE)
mbk2net <- tbl_graph(nodes = mbknodes, edges = mbk2edges, directed = FALSE)
mbk3net <- tbl_graph(nodes = mbknodes, edges = mbk3edges, directed = FALSE)
mbk4net <- tbl_graph(nodes = mbknodes, edges = mbk4edges, directed = FALSE)

#At the end of this section, let's do a little housekeeping and get rid of the datasets we won't be needing anymore:
rm(mbk1, mbk2, mbk3, mbk4, mbkphases, mbk1edges, mbk2edges, mbk3edges, mbk4edges)
```
We have now created four networks which represent four different stages of the Michelsberg culture. We now want to know what changes in these networks and what this might mean for the social networks that exchanged pottery types. Did they change? In what way?
# Visual comparison
It is often a good idea to have a visual exploration of the data in the first place. So just for this, we will make some quick representations of the four graphs.
```{r exploration graphs}
#Because we have to run the same steps on four graphs, we will write some helper functions. This is where R comes into its own. Imagine the number of clicks you would have to do on UCInet or similar to get this!

#first we make a function for thresholding:
thresh.graph <- function(data,x){
  data %>% 
    activate(edges) %>%  # select the edges for transformation
    filter(weight >= x) %>%  # remove all connections below a specified number of shared types
    activate(nodes) %>%  # select nodes for transformation
    filter(!node_is_isolated())  # remove all nodes that do not have a connection
}

#Then we set out our graphics options to shorten the commands
red.weight <- list(
                geom_edge_link(aes(color = weight)),  # color edges according to weight
                geom_node_point(), # add a point for each node
                geom_node_text(aes(label = site_name), color = "red") # add labels to the nodes
                )

#Now we can quickly return the four graphs

mbk1net %>% thresh.graph(4) %>% ggraph(layout = "stress")+ red.weight + labs(title = "Phase 1")
mbk2net %>% thresh.graph(4) %>% ggraph(layout = "stress")+ red.weight + labs(title = "Phase 2")
mbk3net %>% thresh.graph(4) %>% ggraph(layout = "stress")+ red.weight + labs(title = "Phase 3")
mbk4net %>% thresh.graph(4) %>% ggraph(layout = "stress")+ red.weight + labs(title = "Phase 4")

```
Having had a look at the graph, and played around with the thresholds, we think we can see that the two middle Phases are much more connected than phases 1 and 4. But to really find out what changes, we will have to rely on network measures. We will first begin with measures on the whole network, then move on to measures on the nodes and edges.

# Network comparison
We will first look at how the network develops through time in terms of how closed or clustered it is. This means that we are looking at how many triads (connected triangles) there are. We measure this either by transitivity, or by the clustering coefficient.Secondly, we will look at how large the network is and thirdly how wide it is.

There are a great number of measures that might characterise a graph. Here, we will use the following as an example:
* size: how many edges does the graph have?
* order: how many nodes does the graph have?
* mean distance: What is the mean distance between any two nodes?
* edge connectivity : How many edges would have to be removed to make the graph become disconnected?
* clustering coefficient : How closed or clustered is the network (what is its tendency to form closed triangles)?
* density: how densely is the graph connected?


```{r network measures comparison}
# Make a function to get together the measures we want in a table
netmeasures <- function(graph, thresh, phase){
  
  trans <- graph %>%
  thresh.graph(thresh) %>% 
  igraph::transitivity()  # calculate transitivity
  
  dens <- graph %>% 
  thresh.graph(thresh) %>% 
  igraph::graph.density()  # calculate density
  
  graph %>% 
    thresh.graph(thresh) %>% 
    mutate(meandist = graph_mean_dist(),  # calculate mean distance
         size = graph_size(),  # number of edges
         order = graph_order(),  # number of nodes
         connectivity = graph_adhesion()) %>%  # connectivity
    select(meandist, size, order, connectivity) %>%  # select measures
    as_tibble() %>%  # turn into dataframe
    head(1) %>%   # select only first row (as data is repetitive)
    mutate(phase = phase,
           transitivity = trans,
           density = dens) %>%   # add all measures together
    select(phase, everything())  # order phase to the front
}

# Now apply the function
a <- netmeasures(mbk1net, 3, 1)
b <- netmeasures(mbk2net, 3, 2)
c <- netmeasures(mbk3net, 3, 3)
d <- netmeasures(mbk4net, 3, 4)

#make a table of all phases
phase.measures <- bind_rows(a,b,c,d)

#housekeeping
rm(a,b,c,d)

#let's have a look at the table
phase.measures
```
We see that there is quite a distance in size and order between the phases, but that it is never far from one node to another. This is due to the nature of the network: when it is about co-occurrences like this one is, this is usual.
More interesting are the remaining measures, which all correlate to show that Phase 2 is the most closely connected. The high connectivity means that the assemblages were most homogeneous, and that therefore the social contacts between the sites extended to a similar range of objects. While the number of connections increases in Phase 3, so too does the fragility of the network, which might indicate that groups form. On this basis we could say that the Michelsberg culture found its strongest cohesion and most resembled a unified social group during Lüning's phase II. In his Phase III, a fragmentation sets in which continues through Phase IV. I hope that this goes some way to show how different these results are from what we might find out by distribution maps, and also from what was done with this data previously.

# Centralities
Within this dynamic, what were the key sites? This is a question we will ask of measures on the nodes within our networks. We have explained these previously and here we will be using:
* degree centrality
* betweenness centrality
* eigenvector centrality

```{r phase centralities}
#Once again, we wil make a function
centralities <- function(graph, thresh, phase){
  
  graph %>% 
  thresh.graph(thresh) %>% 
  activate(nodes) %>%  # select nodes for transformation
  mutate(degree = centrality_degree(weights = weight),
         betweenness = centrality_betweenness(weights = weight),
         eigen = centrality_eigen(weights = weight),
         phase = phase) %>% 
  as_tibble() %>% 
  select(phase, id, site_name, degree, betweenness, eigen)
}

# Apply the function to each network
p1 <- centralities(mbk1net, 3, 1)
p2 <- centralities(mbk2net, 3, 2)
p3 <- centralities(mbk3net, 3, 3)
p4 <- centralities(mbk4net, 3, 4)
```
These datasets are now the basis for us to have a look at centrality in a couple of ways: by measure, and by site. If we look at them by measure, we can find out what the highest-scoring nodes in each network are.
```{r centralities by measure}
test <- p1 %>% select(id, degree)



centr.ph <- list(p1, p2, p3, p4)



# First, we will take degree centrality and find the highest-scoring
d <- lapply(centr.ph, function(x){ x %>% dplyr::select(id,degree)})
t <- purrr::map(centr.ph, function(x){x %>% select(site_name,degree)}) %>% purrr::reduce(left_join, by = "site_name")
```

